Casual Bandits

### 2016-01-31 causal-bandit/icml-2015/general.tex
modifies definition of m for general case
some explaination of general case, changes to earlier sections to improve flow to general case

If all the variables in $\boldsymbol{X}$ are observable, we can compute the distribution $P(Y,\boldsymbol{X}|do(X_i = j)$ for any intervention $i,j$ from data collected without intervening via the truncated product formula \cite{Pearl2000} 1.37.

\eq {
P(Y,X) = \prod_{k \neq i}\P{x_k|\calP_k(\x)}\P{Y|\calP_Y(\x)}\, \text{  $\forall \,\x$ consistent with $X_i = j$}  
}

@@

We can then marginalize to obtain $P(Y|do(X_i = j))$

\eq {
P(Y|do(X_i = j)) =& \sum_{\x:x_i=j} \prod_{k \neq i}\P{x_k|\calP_k(\x)}\P{Y|\calP_Y(\x)}\\
 =& \sum_{\x:x_i=j} \prod_{k \neq i}\P{x_k|\calP_k(\x)}\P{Y|\x}\\
 =& \sum_{\x:x_i=j} \P{\boldsymbol{X}|do(X_i=j)}\P{Y|\x}
}

@@

>>>>

If all the variables in $\boldsymbol{X}$ are observable, we can compute the distribution $P(Y,\boldsymbol{X}|do(X_i = j)$ for any intervention $i,j$ via the truncated product formula \cite{Pearl2000} 3.11

\eq {
\P{\boldsymbol{x},y|do(X_i = j)} = 
\begin{cases}
\frac{\P{\boldsymbol{x},y}}{\P{x_i=j|\calP_i(\boldsymbol{x})}} & \text{if} x_i = j  \\
0 & \text{otherwise}
\end{cases}
}

@@

Since estimating $\P{Y|do(X_i=j)}$ from observational data requires conditioning on the parents of $X_i$, the variance of our estimators depends on $\P{X_i=j|\calP_i}$, rather than simply being a function of the number of times we observe $X_i=j$ as in the parallel bandit case. For example consider the graph in \cref{fig:causalStructure_confounded}. In this model 
$P(Y|do(X_i = j) = \sum_z P(Y|X_i = j,Z)P(Z)$. Let $P(z=1)=.5$, $P(X_i = 1|z = 1) = 1$, and $P(X_i = 0|z = 0) = 1$ then $P(X_i = 1) = .5$ but we will be unable to estimate $P(Y|do(X_i = 1))$ as we will never get any data for the term $P(Y|X_i = 1, Z = 0)$.

\begin{figure}[h]
\centering
\caption{Causal model for the confounded parallel bandits problem}.\label{fig:causalStructure_confounded}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]

 %nodes

\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, below right=of 2](5){Y};
\node[main node, above=of 3](6){$Z$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (5)
    	(2) edge (5)
    (4) edge (5)
    (6) edge (1) edge (2) edge(4);

\end{tikzpicture}
\end{figure}

For general models, it is also insufficient to only trade off learning about multiple actions by observing against learning individual actions by explicitly playing them. We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.

\eqn {
\label{eq:estimation_transfer}
P(Y_t|do(X_{t,i} = j))= & \nonumber \\
 \sum_{j'}  P(Y_t|do(X_{t,l} & = j'),X_{t,i} = j)P(X_{t,l} = j')
}

Consider the graph in \cref{fig:causalchain}, where each variable deterministically takes the value of its parent,$X_k = X_{k-1}$ for $k\in {N...2}$ and $\P{X_1} = 0$. We can learn the reward for all the arms $do(X_i = 1)$ simultaneously by selecting $do(X_1 = 1)$. If we only considered information transfer from one action to another for the case of the $do()$ action, we would have to explicitly play each of these arms individually to estimate their rewards.

\begin{figure}[h]
\centering
\caption{A causal chain graph}.
\label{fig:causalchain}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}, hidden/.style={empty}]

 %nodes
\node[main node](1){$X_{1}$};
\node[main node, right=of 1](2){$X_{2}$};
\node[hidden, right=of 2](3){$...$};
\node[main node, right=of 3](4){$X_{N}$};
\node[main node, right=of 4](5){Y};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    	(2) edge (3)
    (3) edge (4)
    (4) edge (5);
\end{tikzpicture}
\end{figure}

\todof{From here onwards needs to be replaced by truncated importance sampling version}

@@

\todof{I think the above should be}

\eq {
m = \min \set{m: \sum_{ij} \ind{V_{ij} \geq m} = m}\,.
}

That is, $m$ should be chosen such that the size of the set variances greater than $m$ is $m$. This ensures all the variances of the remaining arms are smaller than $m$. This definition would also seem equivalent to sorting the variances such that $V^{(0)} \geq V^{(1)} ... \geq V^{(2N)}$ and then defining $m = \min m: V^{(m)} \leq m$ which is in turn equivalent to the definition in terms of $q$ for parallel bandits.

### 2016-01-31 causal-bandit/icml-2015/setup.tex
some explaination of general case, changes to earlier sections to improve flow to general case

From a causal perspective the key insight is that the assumed causal structure implies that
the law of $Y_t$ given we intervene to set $X_{t,i} = j$ is the same as the law of $Y_t$ conditioned
on $X_{t,i} = j$. Formally,
\eqn {
\label{eq:observe}
\P{Y_t \leq y|do(X_{t,i} = j)} = \P{Y_t \leq y|do(),X_{t,i} = j}\,.
}
This follows from application of the do-calculus \cite{Pearl2000} to the causal graph for the parallel bandits problem (Fig.~\ref{fig:causalStructure}). It is not the case in general. 
For example if there was a variable $X'$ that caused both $X_{t,i}$ and $Y$ (or $X_{t,i}$ and any other variable $X_{t,l}$), 
that would introduce a backdoor path from $X_{t,i} \rightarrow Y_t$ and we would have to condition on $X'$ to derive the 
interventional distribution of $Y_t$ from the observational one.

\todot{Or just give a reference or hide in the appendix depending on space in the end}
\color{red} Probably should define causal model and back-door rule to properly show this.\color{black}

We can also learn about the reward for intervening on one variable from rounds in which we actually set a different variable.

\eqn {
\label{eq:estimation_transfer}
P(Y_t|do(X_{t,i} = j))= & \nonumber \\
 \sum_{j'}  P(Y_t|do(X_{t,l} & = j'),X_{t,i} = j)P(X_{t,l} = j')
}

\begin{definition}
Let $\boldsymbol{q} \in [0,1]^N$. Then define $q^{(i)}$ to be the $i$th largest value of the set $\set{\min \set{q_j, 1 - q_j} : 1 \leq j \leq N}$.
For example
\eq{
\boldsymbol{q} &= (0,\, 2/3,\, 1/2,\, 1) \\
q^{(1)} &= 0,\,\, q^{(2)} = 0,\,\, q^{(3)} = 1/3,\,\, q^{(4)} = 1/2\,.
}
Define 
\eqn{
\label{def:m}
m(\boldsymbol{q}) = \min\set{1 \leq i \leq N: q^{(i+1)} \geq \frac{1}{i}}\,.
}
\end{definition}

>>>>

From a causal perspective the key insight is that the assumed causal structure implies that
the law of $Y_t$ given we intervene to set $X_{t,i} = j$ is the same as the law of $Y_t$ conditioned
on $X_{t,i} = j$. Formally,
\eqn {
\label{eq:observe}
\P{Y_t \leq y|do(X_{t,i} = j)} = \P{Y_t \leq y|do(),X_{t,i} = j}\,.
}
This follows from application of the do-calculus \cite{Pearl2000} to the causal graph for the parallel bandits problem (\cref{fig:causalStructure}). It is not the case in general. 
For example if there was a variable $X'$ that caused both $X_{t,i}$ and $Y$ (or $X_{t,i}$ and any other variable $X_{t,l}$),that would introduce a backdoor path from $X_{t,i} \rightarrow Y_t$ and we would have to condition on $X'$ to derive the interventional distribution of $Y_t$ from the observational one.

This enables us to estimate the the rewards for multiple interventions simultaneously by observing the values taken by the variables when we select the $do()$ action. Unfortunately interventions that occur with low probability ($\P{X_{t,i} = j}$ is small) suffer from a high approximation error and need to be explored separately, by explicitly making the intervention.

To exploit the causal structure we must optimally balance the collection of observational data against making interventions to learn about low probability events. The problem is more challenging because the probabilities $P(X_i=j)$ are unknown and must also be estimated.

\begin{definition}
Let $V_{ij} = \frac{1}{P(X_i = j)}$, for $i \in \set{1...N}$ and $j \in {0,1}$. 
Define 
\eqn{
\label{def:m}
m(\boldsymbol{V}) = \min \set{m: \sum_{ij} \ind{V_{ij} \geq m} = m}\,.
}
\end{definition}

### 2016-01-31 causal-bandit/icml-2015/simple-regret.tex
More on parallel bandits.
some explaination of general case, changes to earlier sections to improve flow to general case

The main idea of the algorithm is that by making no intervention (doing nothing) it is possible to estimate simultaneously the returns of all interventions.
Unfortunately interventions that occur with low probability ($\P{X_{t,i} = j}$ is small) suffer from a high approximation error and should be
explored separately by actually making the intervention. The algorithm works by optimally balancing the collection of observational data (when
no action is taken) and making interventions to learn about low probability events. The problem is more challenging because $\boldsymbol{q}$ is
unknown and must also be estimated.

\begin{algorithm}[h]
\caption{Causal Best Arm Identification}\label{alg:simple}
\begin{algorithmic}[1]
\STATE {\bf Input:} $T, N$
\FOR{$t \in 1,\ldots,(T - 1) / 2$}
\STATE Choose the action $do()$ and observe $\boldsymbol{X}_t$ and $r_t$
\ENDFOR
\STATE Estimate $\mu$ using observation data:
\eq{
(\forall i,j) \qquad \hat \mu_{i,j} = \frac{\sum_{t=1}^{T/2} \ind{X_{t,i} = j} r_t}{\sum_{t=1}^{T/2} \ind{X_{t,i} = j}}\,.
}
\STATE Compute $\hat q_i = \frac{2}{T} \sum_{t=1}^{T/2} X_{t,i}$
\STATE Compute $\hat s_i = \min\set{\hat q_i, 1 - \hat q_i}$
\STATE Compute $\hat{s'} = sorted(\hat{s}) : \hat{s'}_1 \leq \hat{s'_2} \leq ... \leq \hat{s'}_N$
\STATE Compute $\hat m = \min\set{1 \leq i \leq N : \hat s'_{i+1} \geq \frac{1}{i}}$
\STATE $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
\STATE Compute $A$ as the subset of infrequently observed arms $\{(i,j):i'(i) \leq \hat m, j = \ind{\hat q_{i} \leq \frac{1}{2}} \}$ with $|A| = \hat m$

@@

\begin{definition}
Let $\boldsymbol{q} \in [0,1]^N$. Then define $q^{(i)}$ to be the $i$th largest value of the set $\set{\min \set{q_j, 1 - q_j} : 1 \leq j \leq N}$.
For example
\eq{
\boldsymbol{q} &= (0,\, 2/3,\, 1/2,\, 1) \\
q^{(1)} &= 0,\,\, q^{(2)} = 0,\,\, q^{(3)} = 1/3,\,\, q^{(4)} = 1/2\,.
}
Define 
\eqn{
\label{def:m}
m(\boldsymbol{q}) = \min\set{1 \leq i \leq N: q^{(i+1)} \geq \frac{1}{i}}\,.
}
\end{definition}

>>>>

\begin{algorithm}[h]
\caption{Causal Best Arm Identification}\label{alg:simple}
\begin{algorithmic}[1]
\STATE {\bf Input:} $T, N$
\FOR{$t \in 1,\ldots,(T - 1) / 2$}
\STATE Choose the action $do()$ and observe $\boldsymbol{X}_t$ and $r_t$
\ENDFOR
\STATE Estimate $\mu$ using observation data:
\eq{
(\forall i,j) \qquad \hat \mu_{i,j} = \frac{\sum_{t=1}^{T/2} \ind{X_{t,i} = j} r_t}{\sum_{t=1}^{T/2} \ind{X_{t,i} = j}}\,.
}
\STATE Compute $\hat V_{ij} = \frac{T}{2 \sum_{t=1}^{T/2}\ind{X_{t,i}=j}} $
\STATE Compute $\hat m = m(\hat{\boldsymbol{V}})$
\STATE $i'(i) = $ the index of $\hat s_i$ in $\hat s'$
\STATE Compute $A$ as the subset of infrequently observed arms $\set{(i,j):V_{ij} \geq m}$ with $|A| = \hat m$

@@
